Joint training of classification model and similarity model for low-resource text classification in chatbot

### 2020-07-30 chatbot_classification/aaai_2021/paper.tex
update

% latest version end

>>>>

\section{Methodology}
In this section, we will describe a system of similarity model fused with classification model, named as SFC, in detail. We come up with 2 different approaches for implementation of SFC system, in which hierarchical SFC is a further improvement of 2-stage SFC for better performance. 
\subsection{2-stage SFC with multi-task learning}
The detailed network structure of 2-stage SFC is shown in Fig. \ref{fig:ver.1.3}. It is composed with 2 separate stages in charge of different jobs.
In stage 1, we can use any auxiliary model or tool to provide top-K most related classes for sampling sentence pairs of both positive samples and negative samples for stage 2. Suppose we have a training data set $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of $N$ data points, in which $x_{i}$ is the user input query and $y_{i}$ is a single class label from a class label set of $C$ classes in total. If we want to generate sentence pair data set $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$, where $(x, x')_{j}$ is sentence pair grouped by 2 user input queries and $l_{j}\in\{0, 1\}$ is similarity class label, see Equ. \ref{eq:similarity label}, 
\begin{align}
l_{j}=\begin{cases} \label{eq:similarity label}
0 & y_{s_1}\neq y_{s_2}\\
1 & y_{s_1}=y_{s_2}
\end{cases}
\end{align}
directly from original data set $\mathcal{D}$ without any sampling strategy, then $\mathcal{D'}$ will have $M=N^2$ data points of sentence pairs, which might be too time-consuming for the training process of stage 2 since we often have hundreds of class labels under task-specific chatbot scenario. Although we may only have a few data points for a single class label, $N$ is still quite a big number.

Therefore, we adopt the idea of adversarial negative sampling strategy raised by \cite{bamler2020extreme}. The key idea of this sampling strategy is to train with negative samples that are hard to distinguish from positive samples. Now suppose we have a user input query $x_{i}$ as $s_{1}$ and top-K most related class labels set $\mathcal{C'}=\{c'_{1}, \dots, c'_{K}\}$ found by auxiliary model or tool in stage 1, where $K$ is a hyperparameter set by us that controls candidate class number, we apply the adversarial negative sampling strategy into our sentence pair sampling process by the following 2 steps:

\textbf{1. positive sentence pair sampling}: During the training process, we have to first make sure that the ground truth class label $y_{i}$ for $x_{i}$ is within the candidate class label set $\mathcal{C'}$. If $y_{i}\notin \mathcal{C'}$, we will manually add $y_{i}$ into $\mathcal{C'}$ by replacing $c'_{K}$ with $y_{i}$. Afterwards, we will randomly sample $P$ sentences $\{x'_{0},x'_{1}, \dots, x'_{P}\}$ with the same class label as $y_{i}$ from data set $\mathcal{D}$ to form the set of sentence pairs with positive label $\mathcal{P'}_{i}=\{[(x_{i}, x'_{0}), 1], [(x_{i}, x'_{1}), 1], \dots, [(x_{i}, x'_{P}), 1]\}$, where $P$ is also a hyperparameter set by us that controls the number of sentences we should sample from each class.

\textbf{2. negative sentence pair sampling}: As for negative sentence pairs, we will also randomly sample $P$ sentences for each class in the negative candidate class label set $\mathcal{C'}\backslash \{y_{i}\}$. The class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are the set of most confusing classes comparing to the ground truth ${y_{i}}$, so we assume that the sentence pairs with negative label grouped by user input query $x_{i}$ and sentences with class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are also strong adversarial negative samples that can help enhance the training speed and performance. According to the same method as positive sentence pair sampling, we can obtain the set of sentence pairs with negative label as $\mathcal{N'}_{i}=\{[(x_{i}, x'_{0}), 0], [(x_{i}, x'_{1}), 0], ..., [(x_{i}, x'_{P\cdot (K-1)}), 0]\}$

In this way, we can generate the sentence pair data set $\mathcal{D'}$ for stage 2 based on the top-K class label set $\mathcal{C'}$ provided by stage 1 as $\mathcal{D'}=\{\mathcal{P'}_{i}\cup \mathcal{N'}_{i}\}_{i=1}^{N}$, in which we will have $M=N\cdot K\cdot P$ data points of sentence pairs in total.

After explaining the data sampling principle for sentence pairs and how the 2 stages in SFC system connect with each other, let's go deeper into model structure within these 2 stages respectively to see how it works.

\subsubsection{stage 1: classification model for providing top-K classes}
As for stage 1, our SFC system can adapt to any auxiliary model or even searching tool based on Term Frequency-Inverse
Document Frequency (TF-IDF) retrieval. In this paper, however, we choose to fine-tune RoBERTa on our task-specific data set $\mathcal{D}$ by adding a softmax classifier on top of it, the same as the structure shown in Fig. \ref{Fig.1.sub.1}, for better overall performance. According to our experimental result, RoBERTa can guarantee that approximately 95\% ground truth label be included in top-5 candidate classes with highest probability after being fine-tuned on task-specific data set with top-1 weighted F score at around 0.75. So we use RoBERTa as encoder of classification model in our setting at stage 1 to ensure low limitation for the performance of stage 2.

For single-sentence text classification tasks, given a data point $x_{i}$ with class label $y_{i}$ from data set $\mathcal{D}$, RoBERTa takes the final hidden state $h_{i}$ of the first token [CLS] as the representation for the whole sequence of $x_{i}$. Let's suppose we have a linear layer as ${\Phi}^C_{i}=W^Ch_{i}+b^C$, where $W^C$ and $b^C$ are trainable parameters, the probability score for $x_{i}$ can be calculated as $p^C_{i}=softmax(W^Ch_{i}+b^C)=softmax({\Phi}^C_{i})$. Then, the loss for stage 1 classification model is shown in Equ. \ref{eq:classification loss}, in which the first term $\varPhi_{i,y_{i}}^C$ encourages high probability for prediction of the correct label $y_{i}$.

\begin{equation}
\begin{aligned}
\mathcal{L}^{C}&=\frac{1}{N}\sum_{i=1}^{N}-log(\frac{exp(\varPhi_{i,y_{i}}^C)}{\sum_{j=1}^{C}exp(\varPhi_{i,j}^C)}) \\&=\frac{1}{N}\sum_{i=1}^{N}-\varPhi_{i,y_{i}}^C+log(\sum_{j=1}^{C}exp(\varPhi_{i,j}^C))
\label{eq:classification loss}
\end{aligned}
\end{equation}

\subsubsection{stage 2: sentence-pair model for multi-task learning}
As for stage 2, we also choose RoBERTa as the encoder for sentence pairs, as shown in Fig. \ref{fig:ver.1.3}. Before sending sentence pairs to Roberta, we pack sentence pairs into a single sequence and separate them with a special token [SEP]. Roberta will also use a learned embedding to every token indicating whether it belongs to the first sentence or the second one. Then, we utilize the cross-attention mechanism inside RoBERTa, and then extract semantic relationship between 2 sentences from the output softmax layer, which is placed on top of the last hidden representation of the [CLS] token.

Before we start to fine-tune our sentence-pair model on task-specific dataset, we first fine-tune RoBERTa on Quora dataset\cite{iyer2017first}, which contains 404,290 potential duplicate question pairs, for transfer learning. Then we continue to do multi-task training on sentence-pair dataset $\mathcal{D'}$ sampled from the top-K candidate classes provided by stage 1.

The 2 tasks for stage 2 training process are listed below.
% latest version end
